---
title: "CosMx Human Frontal Cortex Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CosMx Human Frontal Cortex Analysis}
  %\VignetteEngine{knitr::rmarkdown}  
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%"
)
```

# Application of Spacelink to CosMx Human Frontal Cortex Dataset

In this vignette, we demonstrate how to apply **Spacelink** to a large-scale, single-cell resolution spatial transcriptomics dataset. 
Here, we use the **CosMx Human Frontal Cortex** dataset as a representative example of how to handle datasets where the number of cells is large enough to make standard kernel-based methods computationally prohibitive.

We walk through the full Spacelink workflow in two stages: (1) **global tissue-level** SVG detection, 
which tests for spatial variability across all cells regardless of cell type, 
and (2) **cell-type-specific** SVG detection, 
which restricts the analysis to cells of a given type to identify spatially variable patterns within that cell population.

The dataset is publicly available [here](https://brukerspatialbiology.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/human-frontal-cortex-ffpe-dataset/).

## 1. Load Packages

```{r eval=FALSE}
library(spacelink)
library(SpatialExperiment)
library(Seurat)
library(sctransform)
library(ggplot2)
library(viridis)
```

## 2. Load Dataset

We load the CosMx cortex dataset stored as a `SpatialExperiment` object. 
We extract three components needed for downstream analysis:

- **`counts`**: the raw gene expression count matrix (genes × cells).
- **`spatial_coords`**: a matrix of 2D spatial coordinates for each cell, used to model spatial autocorrelation.
- **`cell_type_info`**: a one-hot encoded matrix of cell type assignments, constructed via `model.matrix()` from the `cell_type` column in the object's metadata. Each column corresponds to a cell type, and each row to a cell, with a value of 1 indicating membership. This format is required as input for cell-type-specific analysis in Section 6.

```{r eval=FALSE}
spe <- readRDS("cosmx_cortex_spe.rds")

counts <- spe@assays@data$counts
spatial_coords <- spatialCoords(spe)
cell_type_info <- model.matrix(~ cell_type - 1, data = spe@colData)
colnames(cell_type_info) <- str_remove(colnames(cell_type_info), "cell_type")
```

## 3. Preprocessing

Before running Spacelink, we normalize the raw count matrix using [SCTransform](https://satijalab.org/seurat/reference/sctransform),
which applies a regularized negative binomial regression to correct for sequencing depth 
and other technical variation across cells. 
The normalized values are stored in the `SCT` assay of the resulting Seurat object and extracted for use in Spacelink.

Note that SCTransform can be computationally intensive for large datasets. 
If runtime is a concern, alternative normalization approaches such as log-normalization (`scran` or simple log1p scaling) 
may be substituted.


```{r eval=FALSE}
seurat_res <- CreateSeuratObject(counts = counts)
seurat_norm = SCTransform(seurat_res, vst.flavor = "v2", verbose = FALSE)
normalized_counts <- seurat_norm@assays$SCT$data
```

## 4. Run Global Spacelink

We first run Spacelink at the **global tissue level** to identify genes 
whose expression is spatially variable across the entire tissue, regardless of cell type composition.

Because this CosMx dataset contains a very large number of cells, 
we use the **Spacelink lite** version by setting `lite = TRUE`. 
The lite version approximates the kernel covariance matrix using a **grid-based spatial binning** scheme, 
which substantially reduces memory and computation time while preserving the accuracy of ESV estimates. 
Specifically, rather than computing pairwise distances between all individual cells, 
Spacelink lite aggregates cells into spatial grid bins and computes kernel values between bin centroids.

The default grid size is set to 5 times the minimum nearest-neighbor distance between cells. 
However, for this CosMx dataset, the cells are so densely and uniformly packed that 
the minimum nearest-neighbor distance is very small, 
and the default grid size results in a grid that is nearly as fine as the original data -- providing little computational benefit. 
We therefore manually set `grid_size = 100` (in the same spatial units as the coordinates) 
to achieve a meaningful reduction in the number of effective spatial units.

We demonstrate the analysis on 6 example genes to keep runtime short. 
The two key output columns to interpret are:

- **`padj`**: Benjamini–Hochberg adjusted p-value for the test of spatial autocorrelation. Genes with `padj < 0.05` are considered statistically significant SVGs.
- **`ESV_adj`**: the Effective Spatial Variability value after adjustment, ranging from 0 to 1, reflecting the strength of spatial autocorrelation. Higher values indicate stronger spatial patterning.

Note that both `padj` and `ESV_adj` will differ from a full-gene-set analysis, 
because the BH correction is applied across all tested genes. 
In this example, adjustment is performed only across the 6 selected genes.


```{r eval=FALSE}
gene_list <- c("CNP","EIF5B","HOPX","KRT19","PHB","SHISA5")
global_results <- spacelink(
  normalized_counts=normalized_counts[gene_list,], 
  spatial_coords=spatial_coords, 
  lite=TRUE, 
  grid_size=100)
```


## 5. Visualize Global Spacelink Results

To qualitatively validate the ESV scores, 
we visualize gene expression in spatial coordinates for the genes with the **highest** and **lowest** ESV among the tested genes. 
A gene with a high ESV should display a clearly structured spatial pattern (e.g., layer-specific enrichment), 
while a gene with a low ESV should appear spatially uniform or random. 
Each cell is plotted as a point at its spatial coordinate, colored by raw count expression on a log1p scale.

```{r eval=FALSE}
gene_names <- rownames(global_results)[c(which.max(global_results$ESV_adj),
                                         which.min(global_results$ESV))]

for(gene_name in gene_names){
  df <- as.data.frame(cbind(spatial_coords, expr = counts[gene_name,]))
  colnames(df) <- c('X0', 'X1', 'expr')
  title <- paste0(gene_name, ", ESV:", round(global_results[gene_name,'ESV'],3))
  print(ggplot(df, aes(x = X0, y = X1, color = expr)) +
    geom_point(size = 1) +
    ggtitle(title) + theme_void() + scale_y_reverse() +
    scale_color_gradient(low = "grey97", high = "red", trans = "log1p", name = "counts") +
    theme(plot.title = element_text(face = "italic", size = 16)))
}
```

```{r eval=FALSE}
## Gene expression plots
```


## 6. Run Cell-type-specific Spacelink

Beyond global spatial variability, we may ask whether a gene's spatial pattern emerges specifically within a particular cell type. 
For this, we use `spacelink_ctSVG`, which performs Spacelink restricted to cells belonging to a specified focal cell type. 
When the `cell_type_proportions` argument is supplied as a **binary one-hot encoded matrix** (as constructed in Section 2), 
the function internally detects this and subsets both the expression matrix and spatial coordinates to cells 
where the focal cell type indicator equals 1. 
`spacelink` is then applied only to those cells, 
testing whether the gene is spatially variable within that cell type's spatial distribution.

This approach is directly applicable to single-cell resolution datasets (such as CosMx), 
where each observation corresponds to a single cell with a definitive cell type label. 
The same function also supports spot-based data with continuous cell type proportion estimates (refer to [Example: Visium Human DLPFC](https://hanbyul-lee.github.io/spacelink/articles/example_visium_human_DLPFC.html)).

The `lite` and `grid_size` arguments work identically to the global `spacelink()` function and are passed through to the internal call. Here we apply it to *oligodendrocytes* as the focal cell type.

As before, note that `padj` and `ESV_adj` values are computed relative to the 6 example genes only, 
and will differ from a full-gene-set run.


```{r eval=FALSE}
cell_type_results <- spacelink_ctSVG(
  normalized_counts=normalized_counts[gene_list,], 
  spatial_coords=spatial_coords, 
  cell_type_proportions=cell_type_info,
  focal_cell_type = "oligodendrocyte",
  lite=TRUE, 
  grid_size=100)
```


## 7. Visualize Cell-type-specific Spacelink Results

We visualize the results of cell-type-specific Spacelink to compare 
how a gene's spatial pattern looks globally versus within oligodendrocytes. 
For each of the highest- and lowest-ct-ESV genes, 
we display the full spatial expression map (all cells), 
with the title reporting both the global ESV and the cell-type-specific ESV (ct-ESV) for reference.

We also first plot the **spatial distribution of oligodendrocytes** themselves. 
Since cell-type-specific Spacelink only tests cells of the focal type, 
understanding where those cells are located in the tissue helps interpret 
whether a gene's ct-ESV reflects true within-type spatial variation or is driven by the cell type's own spatial clustering.


```{r eval=FALSE}
gene_names <- rownames(cell_type_results)[c(which.max(cell_type_results$ESV_adj),
                                            which.min(cell_type_results$ESV))]

# Visualize oligodendrocyte proportions
df <- as.data.frame(cbind(spatial_coords, expr = cell_type_proportions[,'oligodendrocyte']))
colnames(df) <- c('X0', 'X1', 'prop')
print(ggplot(df, aes(x = X0, y = X1, color = prop)) +
        geom_point(size = 1) +
        ggtitle("Oligodendrocyte proportions") + theme_void() + scale_y_reverse() +
        scale_color_viridis(option="magma", direction=-1, limits=c(0,1)) +
        theme(plot.title = element_text(face = "italic", size = 16)))

# Visualize gene expression (compare ESV vs ct-ESV)
for(gene_name in gene_names){
  df <- as.data.frame(cbind(spatial_coords, expr = counts[gene_name,]))
  colnames(df) <- c('X0', 'X1', 'expr')
  title <- paste0(gene_name,
                  ", ESV:", round(global_results[gene_name,'ESV'],3),
                  ", ct-ESV:", round(cell_type_results[gene_name,'ESV'],3))
  print(ggplot(df, aes(x = X0, y = X1, color = expr)) +
          geom_point(size = 1) +
          ggtitle(title) + theme_void() + scale_y_reverse() +
          scale_color_gradient(low = "grey97", high = "red", trans = "log1p", name = "counts") +
          theme(plot.title = element_text(face = "italic", size = 16)))
}
```

```{r eval=FALSE}
## Cell type plot for oligodendrocyte
## Gene expression plots
```


## 8. Comparison of Spacelink Lite vs. Original Version

Finally, we benchmark the **lite version** against the **original (full kernel) version** of Spacelink 
in terms of both computational efficiency and similarity of ESV estimates.

Applying the original version to the full CosMx cortex dataset (hundreds of thousands of cells) is computationally infeasible 
due to the O(n²) scaling of the kernel covariance matrix. 
We therefore restrict the comparison to **endothelial cells** only, 
which constitute a small enough subset of the data for the original version to run in reasonable time. 

We then produce two scatter plots:

- **ESV comparison**: Each point represents one gene, with the x-axis showing the ESV from the original version and the y-axis showing the ESV from the lite version. High concordance (points along the diagonal) indicates that the lite approximation preserves the accuracy of the full kernel approach.
- **Runtime comparison**: A bar or scatter plot showing wall-clock time for each version, illustrating the speedup factor achieved by the lite approximation.

The results confirm that the lite version produces ESV estimates highly concordant with the original, while running approximately 2 times faster -- validating its use as the recommended approach for large-scale datasets.

```{r eval=FALSE}
normalized_counts_endothelial <- normalized_counts[,cell_type_info[,'endothelial']==1]
spatial_coords_endothelial <- spatial_coords[cell_type_info[,'endothelial']==1,]

global_results_original <- spacelink(
  normalized_counts = normalized_counts_endothelial,
  spatial_coords = spatial_coords_endothelial,
  lite = FALSE
)

global_results_lite <- spacelink(
  normalized_counts = normalized_counts_endothelial,
  spatial_coords = spatial_coords_endothelial,
  lite = TRUE
)
```

```{r eval=FALSE}
## ESV scatter plot: lite vs. original
## Runtime comparison plot
```
