---
title: "Getting Started with Spacelink"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spacelink}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%"
)
```

The **spacelink** package consists of two main functions: [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) and [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html).

The [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) function performs hypothesis testing at the global tissue level to identify spatially variable genes (SVGs) using an adaptive multi-kernel approach. It also provides **Effective Spatial Variability (ESV)** scores (ranging from 0 to 1), where higher values indicate greater spatial variability in gene expression.

The [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html) function performs hypothesis testing to identify cell-type–specific spatially variable genes (ctSVGs) using a data-driven colocalization gating strategy. It reports **cell-type–specific Effective Spatial Variability (ct-ESV)** scores (ranging from 0 to 1), where higher values indicate greater spatial variability within the focal cell type.

## Installation

Install the development version from GitHub:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("hanbyul-lee/spacelink")
```

```{r}
library(spacelink)
```

## Usage Overview

### 1. Global Analysis

At the global tissue level, [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) requires two primary inputs: a normalized count matrix (`normalized_counts`) and a two-dimensional spatial coordinate matrix (`spatial_coords`).
The normalized count matrix can be obtained, for example, by applying normalization methods such as the [SCTransform](https://satijalab.org/seurat/reference/sctransform) function to raw count data.
Optionally, for large-scale datasets where computational cost is high, setting `lite = TRUE` enables a grid-based approximation of spacelink to accelerate computation.
Parallel computation across genes can be performed by specifying the number of workers (`n_workers`).

```{r eval=FALSE}
global_results <- spacelink(
  normalized_counts = normalized_expression_matrix,  # Genes × Spots
  spatial_coords = coordinates,  # Spots × 2
  lite = FALSE,
  n_workers = 1
)

# View results
head(global_results[order(global_results$ESV_adj, decreasing=T), c("pval","padj","ESV","ESV_adj")])
```

The output includes the following results for each gene:

- `pval` : Combined p-value for testing whether the gene is spatially variable
- `padj` : Benjamini-Hochberg adjusted p-value
- `ESV` : Effective Spatial Variability score (0–1) for prioritizing SVGs; higher values indicate greater spatial variability
- `ESV_adj` : ESV adjusted for non-SVGs (set to 0 if `padj` > 0.05)

### 2. Cell-Type-Specific Analysis

At the cell-type level, [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html) requires four inputs: a normalized count matrix (`normalized_counts`), a two-dimensional spatial coordinate matrix (`spatial_coords`), a cell type proportion matrix (`cell_type_proportions`), and the name of the focal cell type (`focal_cell_type`).
The cell type proportion matrix can be obtained, for example, using methods such as the [RCTD](https://github.com/dmcable/spacexr) function. The `focal_cell_type` must match one of the column names in `cell_type_proportions`.
If all spots are pure (i.e., no cell type mixing), as in single-cell resolution data, the [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) function is internally applied to the subset of spots corresponding to the focal cell type.
Parallelization across genes is supported via the `n_workers` parameter.

```{r eval=FALSE}
cell_type_results <- spacelink_ctSVG(
  normalized_counts = normalized_expression_matrix,  # Genes × Spots
  spatial_coords = coordinates,  # Spots × 2
  cell_type_proportions = cell_type_proportions,  # Spots × Cell-types
  focal_cell_type = cell_type_name,  # Cell type name to test
  n_workers = 1
)

# View results
head(cell_type_results[order(cell_type_results$ESV_adj, decreasing=T), c("pval","padj","ESV","ESV_adj")])
```

The output includes the following for each gene:

- `pval` : Combined p-value for testing spatial variability within the focal cell type
- `padj` : Benjamini-Hochberg adjusted p-value
- `ESV` : Cell-type–specific Effective Spatial Variability (ct-ESV) score (0–1) for prioritizing ctSVGs; higher values indicate greater spatial variability within the focal cell type
- `ESV_adj` : ESV adjusted for non-ctSVGs (set to 0 if `padj` > 0.05)