---
title: "Spacelink Package Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spacelink package overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%"
)
```

The **spacelink** package consists of two main functions: 

- [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html): Performs hypothesis testing at the *global tissue level* to identify spatially variable genes (SVGs) using an adaptive multi-kernel approach. It also provides **Effective Spatial Variability (ESV)** scores (0-1), where higher values indicate stronger spatial variability in gene expression.
- [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html): Performs hypothesis testing to identify *cell-type–specific* spatially variable genes (ctSVGs) using a data-driven colocalization gating strategy. It reports **cell-type–specific Effective Spatial Variability (ct-ESV)** scores (ranging from 0 to 1), where higher values indicate stronger spatial variability within the focal cell type.

## 1. [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) - Global Analysis

### Inputs

[`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) requires two primary inputs: 

- `normalized_counts`: Normalized count matrix (genes x spots). This can be generated, for example, by applying normalization methods such as [SCTransform](https://satijalab.org/seurat/reference/sctransform) to raw count data. See the [CosMx](https://hanbyul-lee.github.io/spacelink/articles/illustration_on_cosmx_data.html) and [Visium](https://hanbyul-lee.github.io/spacelink/articles/illustration_on_visium_data.html) application examples for normalization details.
- `spatial_coords`: Two-dimensional spatial coordinate matrix (spots x 2).

Optionally, you can provide:

- `lite`: Logical variable. If `TRUE`, a grid-based approximation is used to speed up computation for large datasets. Default is `FALSE`. For very large datasets, we recommend setting `lite = TRUE`.
- `grid_size`: Grid cell size when `lite = TRUE`. If `NULL`, It is set automatically to 5 times the minimum nearest-neighbor distance. Default is `NULL`. We recommend adjusting the grid size to suit your dataset.
- `n_workers`: Number of workers for parallel computation across genes.

### Outputs

- `pval` : Combined p-value testing whether each gene is spatially variable.
- `padj` : Benjamini-Hochberg adjusted p-value.
- `ESV` : Effective Spatial Variability score (0–1) for prioritizing SVGs; higher values indicate stronger spatial variability.
- `ESV_adj` : ESV adjusted for non-SVGs (set to 0 if `padj` > 0.05).

### Code example

```{r eval=FALSE}
global_results <- spacelink(
  normalized_counts = normalized_expression_matrix,
  spatial_coords = coordinates,
  lite = FALSE,
  n_workers = 1
)

# View results
head(global_results[order(global_results$ESV_adj, decreasing=T), c("pval","padj","ESV","ESV_adj")])
```



## 2. [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html) - Cell-Type-Specific Analysis

### Inputs

[`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html) requires four primary inputs:

- `normalized_counts`: Normalized count matrix (as above).
- `spatial_coords`: Two-dimensional spatial coordinate matrix (spots x 2).
- `cell_type_proportions`: Cell type proportion matrix (spots x cell-types), e.g., estimated with [RCTD](https://github.com/dmcable/spacexr). If all spots contain only one cell type (no mixing), [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) is applied internally to the subset of spots corresponding to the focal cell type.
- `focal_cell_type`: Name of the focal cell type; must match one of the columns in `cell_type_proportions`.

Optionally, you can provide:

- `lite`: Logical variable, applicable only when all spots are pure (single-cell resolution). If `TRUE`, a grid-based approximation is used to speed up computation for large datasets. Default is `FALSE`. For very large numbers of cells, we recommend setting `lite = TRUE`.
- `grid_size`: Grid cell size when `lite = TRUE`. If `NULL`, set automatically to 5 times the minimum nearest-neighbor distance. Default is `NULL`. Adjust to suit your dataset.
- `n_workers`: Number of workers for parallel computation across genes.

### Outputs

- `pval` : Combined p-value testing spatial variability within the focal cell type.
- `padj` : Benjamini-Hochberg adjusted p-value.
- `ESV` : Cell-type–specific Effective Spatial Variability (ct-ESV) score (0–1) for prioritizing ctSVGs; higher values indicate stronger spatial variability within the focal cell type.
- `ESV_adj` : ESV adjusted for non-ctSVGs (set to 0 if `padj` > 0.05).

### Code example

```{r eval=FALSE}
cell_type_results <- spacelink_ctSVG(
  normalized_counts = normalized_expression_matrix,
  spatial_coords = coordinates,
  cell_type_proportions = cell_type_proportions,
  focal_cell_type = cell_type_name,
  lite = FALSE,
  n_workers = 1
)

# View results
head(cell_type_results[order(cell_type_results$ESV_adj, decreasing=T), c("pval","padj","ESV","ESV_adj")])
```