---
title: "Spacelink Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spacelink Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%"
)
```


Here, we demonstrate the workflow for spatial gene variability analysis at the *global tissue level* using the [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) function through a simple simulation example.


## 1. Simulate Data

Here, gene expression data ($x$) are simulated from the following simple Gaussian process model:

$y \sim MVN(0, \alpha K_l + I),$

$x \sim Poisson(\exp(y))$

where $K_l$ is an exponential covariance kernel with length scale $l$.

We simulate a total of 6 spatially variable genes (SVGs) and 4 non-SVGs.
For the first 3 genes (`Gene_1`, `Gene_2`, `Gene_3`), the length scale $l$ is set to 0.5, and the signal-to-noise ratio $\alpha$ is set to 0.5, 1, and 2, respectively.
For the next 3 genes (`Gene_4`, `Gene_5`, `Gene_6`), the length scale $l$ is set to 2, and the signal-to-noise ratio $\alpha$ is again set to 0.5, 1, and 2, respectively.
The non-SVGs (`Gene_7`, `Gene_8`, `Gene_9`, `Gene_10`) are simulated with $\alpha=0$.

SVGs are expected to exhibit higher spatial variability as the length scale and signal-to-noise ratio increase.

```{r eval=FALSE}
set.seed(123)
n_spots <- 500
n_genes <- 10

# Obtain spatial coordinates
coords <- expand.grid(
  x = seq(0, 10, length.out = 25),
  y = seq(0, 10, length.out = 20)
)

expr_data <- matrix(nrow = n_genes, ncol = n_spots)
rownames(expr_data) <- paste0("Gene_", 1:n_genes)
D <- as.matrix(dist(coords))

# Simulate Gene_1, Gene_2, Gene_3
lengthscale = 0.5
snr = c(0.5, 1, 2)
K <- exp(-D/lengthscale); chol_K <- chol(K)
for (i in 1:3) {
  expr_data[i, ] <- rnorm(n_spots) + snr[i] * rnorm(n_spots) %*% chol_K
}

# Simulate Gene_4, Gene_5, Gene_6
lengthscale = 2
snr = c(0.5, 1, 2)
K <- exp(-D/lengthscale); chol_K <- chol(K)
for (i in 1:3) {
  expr_data[i+3, ] <- rnorm(n_spots) + snr[i] * rnorm(n_spots) %*% chol_K
}

# Simulate Gene_7, Gene_8, Gene_9, Gene_10
for (i in 7:10) {
  expr_data[i, ] <- rnorm(n_spots)
}

expr_data <- apply(exp(expr_data), 1:2, function(x)rpois(1,x))
expr_data[,1:10]
```

```{r eval=FALSE}
##         [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10]
## Gene_1     1    0    5    1    0    4    1    0    0     0
## Gene_2     0    0    0    1    0   29    5    9    2     0
## Gene_3     1    1    0    0    5    0    1    6    1     4
## Gene_4     2    1    1    1    8    0    3    0    2     0
## Gene_5     0    2    6    1    1   27    9    7    1     3
## Gene_6     0    0    0    0    0    0    6    3    0     0
## Gene_7     1    4    2    1    2    1    2    5    1     0
## Gene_8     1    1    2    3    0    4    2    1    9     3
## Gene_9     0    0    6    3    7    0    0    4    1     1
## Gene_10    2    0    2    0    1    0    1    0    3     0
```


## 2. Preprocessing

Since [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) requires log-transformed normalized input data, we first normalize the count data (`expr_data`) before running the method.
In this example, we apply a simple log normalization, as the data are simple.
For more sophisticated normalization approaches that also correct for differences in sequencing depth across spots, please refer to the [CosMx](https://hanbyul-lee.github.io/spacelink/articles/illustration_on_cosmx_data.html) and [Visium](https://hanbyul-lee.github.io/spacelink/articles/illustration_on_visium_data.html) application examples.

```{r eval=FALSE}
norm_data <- log1p(expr_data)/log(2)
norm_data[,1:10]
```

```{r eval=FALSE}
##             [,1]     [,2]     [,3] [,4]     [,5]     [,6]     [,7]     [,8]     [,9]    [,10]
## Gene_1  1.000000 0.000000 2.584963    1 0.000000 2.321928 1.000000 0.000000 0.000000 0.000000
## Gene_2  0.000000 0.000000 0.000000    1 0.000000 4.906891 2.584963 3.321928 1.584963 0.000000
## Gene_3  1.000000 1.000000 0.000000    0 2.584963 0.000000 1.000000 2.807355 1.000000 2.321928
## Gene_4  1.584963 1.000000 1.000000    1 3.169925 0.000000 2.000000 0.000000 1.584963 0.000000
## Gene_5  0.000000 1.584963 2.807355    1 1.000000 4.807355 3.321928 3.000000 1.000000 2.000000
## Gene_6  0.000000 0.000000 0.000000    0 0.000000 0.000000 2.807355 2.000000 0.000000 0.000000
## Gene_7  1.000000 2.321928 1.584963    1 1.584963 1.000000 1.584963 2.584963 1.000000 0.000000
## Gene_8  1.000000 1.000000 1.584963    2 0.000000 2.321928 1.584963 1.000000 3.321928 2.000000
## Gene_9  0.000000 0.000000 2.807355    2 3.000000 0.000000 0.000000 2.321928 1.000000 1.000000
## Gene_10 1.584963 0.000000 1.584963    0 1.000000 0.000000 1.000000 0.000000 2.000000 0.000000
```


## 3. Run Spacelink

We now run [`spacelink`](https://hanbyul-lee.github.io/spacelink/reference/spacelink.html) using the log-transformed normalized data (`norm_data`) and the spatial coordinates (`coords`).

```{r eval=FALSE}
library(spacelink)

results <- spacelink(normalized_counts = norm_data, spatial_coords = coords)
results[, c("padj", "ESV")]
```

```{r eval=FALSE}
##                 padj         ESV
## Gene_1  2.388026e-01 0.081564906
## Gene_2  3.376990e-02 0.171733914
## Gene_3  1.378931e-08 0.467050678
## Gene_4  4.138725e-02 0.124286163
## Gene_5  4.901449e-20 0.418612352
## Gene_6  5.660031e-47 0.786031823
## Gene_7  7.575684e-01 0.000000000
## Gene_8  5.102295e-01 0.014419457
## Gene_9  7.999929e-01 0.000000000
## Gene_10 5.102295e-01 0.005973537
```


## 4. Analyze Results

We can identify which genes are detected as SVGs based on the hypothesis testing results as follows:

```{r eval=FALSE}
svg_results <- results[results$padj < 0.05,]
rownames(svg_results)
```

```{r eval=FALSE}
## [1] "Gene_2" "Gene_3" "Gene_4" "Gene_5" "Gene_6"
```

The simulated SVGs -- `Gene_2`, `Gene_3`, `Gene_4`, `Gene_5`, and `Gene_6` -- are successfully identified as SVGs, except for `Gene_1`, which was simulated under low length scale and low signal-to-noise ratio.

Next, among the identified SVGs, we can assess which genes exhibit the highest spatial variability using the **Effective Spatial Variability (ESV)** score:

```{r eval=FALSE}
svg_results = svg_results[order(svg_results$ESV, decreasing=T),]
svg_results$rank = 1:nrow(svg_results)
svg_results[,c('rank','ESV')]
```

```{r eval=FALSE}
##        rank       ESV
## Gene_6    1 0.7860318
## Gene_3    2 0.4670507
## Gene_5    3 0.4186124
## Gene_2    4 0.1717339
## Gene_4    5 0.1242862
```

`Gene_6`, which was simulated with the highest length scale and signal-to-noise ratio, shows the highest ESV value and is ranked first.
