---
title: "Spacelink ctSVG Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spacelink ctSVG Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%"
)
```

Here, we demonstrate the spatial gene variability analysis at the *cell-type–specific level* using the [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html) function through a simple simulation example.


## 1. Simulate Data

Here, gene expression data ($x$) are simulated from the following Gaussian process model:

$y \sim MVN(0, \alpha\cdot \Pi_r K_l \Pi_r + I)$

$x \sim Poisson(\exp(y))$

where $\Pi_r$ is a diagonal matrix of the proportions of cell type $r$, and $K_l$ is an exponential covariance kernel with length scale $l$.

We consider there are two cell types (`Cell_type_1`, `Cell_type_2`).
We simulate a total of six cell-type-1-specific spatially variable genes (ctSVGs), two cell-type-2-specific ctSVGs and two non-SVGs.
For the first three cell-type-1 ctSVGs (`Gene_1`, `Gene_2`, `Gene_3`), the length scale $l$ is set to 0.5, and the signal-to-noise ratio $\alpha$ is set to 1, 2, and 5, respectively.
For the next three cell-type-1 ctSVGs (`Gene_4`, `Gene_5`, `Gene_6`), the length scale $l$ is set to 2, and the signal-to-noise ratio $\alpha$ is again set to 1, 2, and 5, respectively.
For the cell-type-2 ctSVGs (`Gene_7`, `Gene_8`), the length scale $l$ is set to 1, and the signal-to-noise ratio $\alpha$ is again set to 1 and 5, respectively.
The non-SVGs (`Gene_9`, `Gene_10`) are simulated with $\alpha=0$.

ctSVGs are expected to exhibit higher spatial variability as the length scale and signal-to-noise ratio increase.

```{r eval=FALSE}
set.seed(123)
n_spots    <- 500
n_genes    <- 10
n_celltypes <- 2

# Obtain spatial coordinates
coords <- expand.grid(
  x = seq(0, 10, length.out = 25),
  y = seq(0, 10, length.out = 20)
)

# Obtain cell type proportions
cell_type_props <- matrix(nrow = n_spots, ncol = n_celltypes)
colnames(cell_type_props) <- paste0("Cell_type_", 1:n_celltypes)
cell_type_props[1:(n_spots/2), 'Cell_type_1'] <- runif(n_spots / 2, 0, 0.5)
cell_type_props[(n_spots/2+1):n_spots, 'Cell_type_1'] <- runif(n_spots / 2, 0.5, 1)
cell_type_props[, 'Cell_type_2'] <- 1 - cell_type_props[, 'Cell_type_1']

expr_data <- matrix(nrow = n_genes, ncol = n_spots)
rownames(expr_data) <- paste0("Gene_", 1:n_genes)
D <- as.matrix(dist(coords))

# Simulate Gene_1, Gene_2, Gene_3
lengthscale = 0.5
snr = c(1, 2, 5)
K <- exp(-D/lengthscale)
Sigma <- chol(K) %*% diag(cell_type_props[, "Cell_type_1"])
for (i in 1:3) {
  expr_data[i, ] <- rnorm(n_spots) + snr[i] * rnorm(n_spots) %*% Sigma
}

# Simulate Gene_4, Gene_5, Gene_6
lengthscale = 2
snr = c(1, 2, 5)
K <- exp(-D/lengthscale)
Sigma <- chol(K) %*% diag(cell_type_props[, "Cell_type_1"])
for (i in 1:3) {
  expr_data[i+3, ] <- rnorm(n_spots) + snr[i] * rnorm(n_spots) %*% Sigma
}

# Simulate Gene_7, Gene_8
lengthscale = 1
snr = c(1, 5)
K <- exp(-D/lengthscale)
Sigma <- chol(K) %*% diag(cell_type_props[, "Cell_type_2"])
for (i in 1:2) {
  expr_data[i+6, ] <- rnorm(n_spots) + snr[i] * rnorm(n_spots) %*% Sigma
}

# Simulate Gene_9, Gene_10
for (i in 9:10) {
  expr_data[i, ] <- rnorm(n_spots)
}

expr_data <- apply(exp(expr_data), 1:2, function(x)rpois(1,x))
expr_data[,1:10]
```

```{r eval=FALSE}
##         [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10]
## Gene_1     0    1    0    0    2    0    4    1    0     0
## Gene_2     3    4    1    0    2    0    0    1    0     0
## Gene_3     0    0    0    5   12    1    1    2    3     0
## Gene_4     0    1    4    1    1    7    8    2    0     0
## Gene_5     4    1    1    0    1    0    2    3    0     2
## Gene_6     1   11    0    1   12    5    6  203    3     2
## Gene_7     2    0    2    0    0    0    0    5    0     7
## Gene_8    70    2 2281   66    0    1    1    2    1     3
## Gene_9     0    1    0    0    1    0    2    2    0     1
## Gene_10    3    0    0    2    1   12    3    4    0     2
```


## 2. Preprocessing

Since [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html) requires log-transformed normalized input data, we first normalize the count data (`expr_data`) before running the method.
In this example, we apply a simple log normalization, as the data are simple.
For more sophisticated normalization approaches that also correct for differences in sequencing depth across spots, please refer to the [CosMx](https://hanbyul-lee.github.io/spacelink/articles/illustration_on_cosmx_data.html) and [Visium](https://hanbyul-lee.github.io/spacelink/articles/illustration_on_visium_data.html) application examples.

```{r eval=FALSE}
norm_data <- log1p(expr_data)/log(2)
norm_data[,1:10]
```

```{r eval=FALSE}
##             [,1]     [,2]      [,3]     [,4]     [,5]     [,6]     [,7]     [,8] [,9]    [,10]
## Gene_1  0.000000 1.000000  0.000000 0.000000 1.584963 0.000000 2.321928 1.000000    0 0.000000
## Gene_2  2.000000 2.321928  1.000000 0.000000 1.584963 0.000000 0.000000 1.000000    0 0.000000
## Gene_3  0.000000 0.000000  0.000000 2.584963 3.700440 1.000000 1.000000 1.584963    2 0.000000
## Gene_4  0.000000 1.000000  2.321928 1.000000 1.000000 3.000000 3.169925 1.584963    0 0.000000
## Gene_5  2.321928 1.000000  1.000000 0.000000 1.000000 0.000000 1.584963 2.000000    0 1.584963
## Gene_6  1.000000 3.584963  0.000000 1.000000 3.700440 2.584963 2.807355 7.672425    2 1.584963
## Gene_7  1.584963 0.000000  1.584963 0.000000 0.000000 0.000000 0.000000 2.584963    0 3.000000
## Gene_8  6.149747 1.584963 11.156083 6.066089 0.000000 1.000000 1.000000 1.584963    1 2.000000
## Gene_9  0.000000 1.000000  0.000000 0.000000 1.000000 0.000000 1.584963 1.584963    0 1.000000
## Gene_10 2.000000 0.000000  0.000000 1.584963 1.000000 3.700440 2.000000 2.321928    0 1.584963
```


## 3. Run Spacelink (ctSVG)

We now run [`spacelink_ctSVG`](https://hanbyul-lee.github.io/spacelink/reference/spacelink_ctSVG.html) using the log-transformed normalized data (`norm_data`), the spatial coordinates (`coords`), and the cell type proportions (`cell_type_props`).

Here, we aim to identify the cell-type-1-specific ctSVGs. Therefore, we set `focal_cell_type` to the corresponding column name, `Cell_type_1`.

```{r eval=FALSE}
library(spacelink)

results <- spacelink_ctSVG(normalized_counts = norm_data, spatial_coords = coords, cell_type_proportions = cell_type_props, focal_cell_type = "Cell_type_1")
results[, c("padj", "ESV")]
```

```{r eval=FALSE}
##                 padj          ESV
## Gene_1  7.135920e-04 3.718855e-01
## Gene_2  2.093598e-08 5.074171e-01
## Gene_3  1.008101e-21 5.922732e-01
## Gene_4  2.839529e-02 2.448116e-01
## Gene_5  5.597474e-04 2.987805e-01
## Gene_6  2.072716e-87 7.598290e-01
## Gene_7  1.079779e-01 1.801354e-01
## Gene_8  9.464224e-01 6.696993e-06
## Gene_9  9.464224e-01 4.681218e-06
## Gene_10 4.884490e-01 4.553093e-02
```


## 4. Analyze Results

We can identify which genes are detected as cell-type-1-specific ctSVGs based on the hypothesis testing results as follows:

```{r eval=FALSE}
ctsvg_results <- results[results$padj < 0.05,]
rownames(ctsvg_results)
```

```{r eval=FALSE}
## [1] "Gene_1" "Gene_2" "Gene_3" "Gene_4" "Gene_5" "Gene_6"
```

All the simulated cell-type-1-specific ctSVGs -- `Gene_1`, `Gene_2`, `Gene_3`, `Gene_4`, `Gene_5`, and `Gene_6` -- are successfully identified.

Next, among the identified cell-type-1-specific ctSVGs, we can assess which genes exhibit the highest spatial variability within the cell type using the **cell-type–specific Effective Spatial Variability (ct-ESV)** score:

```{r eval=FALSE}
ctsvg_results = ctsvg_results[order(ctsvg_results$ESV, decreasing=T),]
ctsvg_results$rank = 1:nrow(ctsvg_results)
ctsvg_results[,c('rank','ESV')]
```

```{r eval=FALSE}
##        rank       ESV
## Gene_6    1 0.7598290
## Gene_3    2 0.5922732
## Gene_2    3 0.5074171
## Gene_1    4 0.3718855
## Gene_5    5 0.2987805
## Gene_4    6 0.2448116
```

`Gene_6`, which was simulated with the highest length scale and signal-to-noise ratio, shows the highest ct-ESV value and is ranked first.
